The Design, Semantics, and Implementation of the Synchronous Language \CEU

### 2015-10-13 paper.tex
(*) submitted

Despite its success and influence, Esterel has an overly complex semantics that 
requires ingenious static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems, with extensive coverage in 
the 
literature~\cite{esterel.constructive,esterel.d1,esterel.d2,esterel.d3,esterel.d4,esterel.d5,esterel.d6,esterel.d7,esterel.d8}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Furthermore, it also affects the capacity of the programmer's understanding, 
which, after all, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of loose and 
non-deterministic execution for intra-reaction statements, which prevents 
threads to share memory and interact with side-effect and stateful system calls 
appropriately.
%
For instance, although Esterel supports orthogonal abortion of 
threads~\cite{esterel.preemption}, it doesn't offer a finalization mechanism to 
release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward implementation targeting resource-constrained systems.
%
The list that follows summarizes the semantic peculiarities of our design:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provide a limited form 
of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution holding 
external resources.
%
\item First-class synchronized timers.
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
(Both use the \emph{ATmega328} microcontroller with 32 Kbytes of FLASH and 2 
Kbytes of SRAM.)
}, showing that the peculiarities in the semantics of \CEU does not pose 
practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, a simple application that blinks three LEDs periodically 
occupies less than 100 bytes and can be transmitted over the air in a couple of 
packets.

@@

\begin{figure}[b]
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (ln. 3--6), in parallel 
with the caller (ln. 8--11).
\label{lst.sub}
}
\end{figure}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU has a different take, posing a tradeoff that an \code{await} is only valid 
for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction (known as \emph{schizophrenic 
statements}~\cite{esterel.schizo,esterel.schizo2}).

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
For instance, sometimes we need to execute a block of code immediately, and 
then, periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with schizophrenic statements.

\begin{figure}[t]
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} restriction by 
post-fixing the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

The unsafe example of Figure~\ref{lst.fin}.a does not compile because \CEU 
tracks the interaction of \code{par/or} compositions with local variables and 
stateful output events calls in order to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
\CEU enforces a \emph{finalization} clause to accompany the output request.
The code in Figure~\ref{lst.fin}.b properly cancels the packet transmission 
when the block of \code{buffer} goes out of scope, i.e., the finalization 
clause (ln. 18) executes automatically on external abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the peculiarities of \CEU, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and system calls (which behave like in conventional 
imperative languages).

@@

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction eventually blocks in $awaiting$ and $emitting$ expressions in 
parallel trails.
%
If all trails hangs only in $awaiting$ expressions, it means that the program 
cannot advance in the current reaction.
%
However, $emitting$ expressions are pending in lower stack indexes and should 
eventually resume in the ongoing reaction (see rule \textbf{pop}).
%
Therefore, we define another rule that behaves as \textbf{rule-inner} 
(presented above) if the program can advance, and, otherwise, pops the stack to 
resume the lower level:
%
$$
\frac
    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
%   -----------------------------------------------------------
    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
%
\5\5\5
%
\frac
    { \DS isBlocked(n,\1s:S,\1p) }
%   -----------------------------------------------------------
    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
\5\5\5\textbf{(rule-outer)}
$$
%
To describe a \emph{reaction} in \CEU, i.e., how a program behaves in reaction 
to a single external event, we use the reflexive transitive closure of 
\textbf{rule-outer}:
%
$$
    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
$$
%
Finally, to describe the complete execution of a program, we trigger multiple 
``invocations'' of reaction chains in sequence:
%
\begin{align*}
\LL [e1], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [  ], p' \RR
\\
\LL [e2], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
\LL [e3], p'' \RR
    & \xRightarrow[~~3~~]{*}
\LL [  ], p''' \RR
\\
& ...
\end{align*}
%
Each invocation starts with an external event at the top of the stack and 
finishes with a modified program and an empty stack.
After each invocation, we increment the sequence number.

@@

\begin{comment}

There are also some restriction on valid programs in \CEU, which the semantic 
allows:
- loops
- awaits in fins
- local scopes

handled in separate, in a parsing phase that is not related to the execution

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The semantic rules are continuously applied (note the `$*$' klenee operator) 
until
consider only the top of the stack and continuously apply transformations to 
the program until it blocks and no rules can be applied.
The \emph{isBlocked} predicate in Figure~\ref{fig.isBlocked} identifies a 
blocked program based on its structure and event at the top of the stack.
xxx
A program becomes blocked when all parallel branches are hanged in $awaiting$, 
$stacked$, and/or $emitting$ primitives, as defined in 
Figure~\ref{fig:isBlocked}.
an \code{awaiting} is unblocked only if its event matches the top of the stack
the \code{emitting} primitive only proceeds once its stack level is restored.

\begin{itemize}
\item The program is awaiting in all trails, i.e., function $pop$ returns 
$(0,\{\})$.
\item The program terminates, i.e., the small-step rules transform the whole 
program into a $nop$.
\end{itemize}
%
In Section~\ref{XXX} we show that by imposing syntactic restrictions to 
programs, reaction chains always reach one of these conditions in a finite 
number of steps, meaning that reactions to the environment always execute in 
bounded time.

To be compliant with the reactive nature of \CEU, we assume that all programs 
start awaiting the main event ``$\$$'', which is emitted once by the 
environment on startup, i.e., $(i,E)=(1,\{\$\})$ for the very first big step.

As briefly introduced, small-step rules continuously apply transformations to 
unblocked trails.
A program becomes blocked when all parallel branches are hanged in $awaiting$, 
$stacked$, and/or $emitting$ primitives, as defined in 
Figure~\ref{fig:isBlocked}.

All small-step rules are associated with the current (deepest) stack depth 
level $i$ acquired from the previous big step.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{comment}

@@

In the sections that follow, we discuss the implementation insights related to 
the peculiarities of \CEU:
the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

@@

    <...>   // other events
}
\end{lstlisting}
%\rule{14cm}{0.37pt}
\caption{
%{\small %\textmd{
The \emph{TinyOS} binding for \CEU.
This platform-dependent template includes the \C files generated from the 
original application in \CEU (\code{ceu\_app.h} and \code{ceu\_app.c}) for the 
\emph{Final generation} phase of Figure~\ref{fig.impl}.
%}%}
\label{lst.impl.tinyos}
}
\end{figure}

@@

A number of virtual machines have been proposed for embedded systems.
%
The \emph{Sun SPOT} platform with the \emph{Squawk JVM} brings Java for the 
embedded domain~\cite{wsn.sunspot}, but requires a much powerful hardware.%
\footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 
KBytes of SRAM.}
%
\emph{Darjeeling}~\cite{wsn.darjeeling} and \emph{TakaTuka}~\cite{wsn.takatuka} 
are complete \emph{JVMs} targeting constrained embedded systems with support 
for multithreading and garbage collection.
%
Java has antagonistic design choices in comparison to \CEU:
it does not impose static bounds on memory usage and execution time, and 
provides preemptive multithreading with synchronization primitives for 
accessing shared memory.
%
Plummer et al.~\cite{esterel.vm} propose a Esterel-based \VM with similar 
design choices to our work.
To reduce code size, the \VM has a specialized instruction set to deal with 
events and concurrency constructs that are particular to Esterel.
However, the proposed \VM is only a proof of concept, with no support for 
arithmetic operations, external system calls, or remote reprogramming.

>>>>

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems.
Both problems have an extensive coverage and debate in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Furthermore, it also affects the programmer's understanding about the code, 
which, after all, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of the loose and 
non-deterministic execution for intra-reaction statements, which prevents 
threads to share memory and interact with side-effect and stateful system calls 
appropriately.
%
For instance, although Esterel supports orthogonal abortion of 
threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward implementation targeting resource-constrained systems.
%
The list that follows summarizes the semantic peculiarities of our design:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provides a limited form 
of coroutines.
%
\item Static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item Finalization mechanism for safe abortion of lines of execution holding 
external resources.
%
\item First-class synchronized timers.
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing that the peculiarities in the semantics of \CEU does not pose 
practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, applications have a bytecode footprint in the order of 
hundreds of bytes can be transmitted over the air in a few 
packets~\cite{ceu.terra}.

@@

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (ln. 3--6), in parallel 
with the caller (ln. 8--11).
\label{lst.sub}
}
\end{figure}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU has a different take, posing a tradeoff that an \code{await} is only valid 
for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction (known as \emph{schizophrenic 
statements}~\cite{esterel.constructive}).

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
For instance, sometimes we need to execute a block of code immediately, and 
then, periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
%
On the one hand, we transfer the burden of dealing with these specific corner 
cases to the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

\begin{figure}[t]
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

The unsafe example of Figure~\ref{lst.fin}.a does not compile because \CEU 
tracks the interaction of \code{par/or} compositions with pointers to local 
variables and stateful output events calls in order to preserve safe abortion 
of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
\CEU enforces a \emph{finalization} clause to accompany the output request.
The code in Figure~\ref{lst.fin}.b properly cancels the packet transmission 
when the block of \code{buffer} goes out of scope, i.e., the finalization 
clause (ln. 18) executes automatically on external abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel highlighting the peculiarities of \CEU, 
in particular the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and system calls (which behave like in conventional 
imperative languages).

@@

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction eventually blocks in $awaiting$ and $emitting$ expressions in 
parallel trails.
%
If all trails hangs only in $awaiting$ expressions, it means that the program 
cannot advance in the current reaction.
%
However, $emitting$ expressions are pending in lower stack indexes and should 
eventually resume in the ongoing reaction (see rule \textbf{pop}).
%
Therefore, we define another rule that behaves as \textbf{rule-inner} 
(presented above) if the program can advance, and, otherwise, pops the stack to 
resume the lower level:
%
$$
\frac
    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
%   -----------------------------------------------------------
    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
%
\5\5\5
%
\frac
    { \DS isBlocked(n,\1s:S,\1p) }
%   -----------------------------------------------------------
    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
\5\5\5\textbf{(rule-outer)}
$$
%
To describe a \emph{reaction} in \CEU, i.e., how a program behaves in reaction 
to a single external event, we use the reflexive transitive closure of 
\textbf{rule-outer}:
%
$$
    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
$$
%
Finally, to describe the complete execution of a program, we trigger multiple 
``invocations'' of reactions in sequence:
%
\begin{align*}
\LL [e1], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [  ], p' \RR
\\
\LL [e2], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
\LL [e3], p'' \RR
    & \xRightarrow[~~3~~]{*}
\LL [  ], p''' \RR
\\
& ...
\end{align*}
%
Each invocation starts with the occurring external event at the top of the 
stack and finishes with a modified program and an empty stack.
After each invocation, we increment the sequence number.

@@

@@

In the sections that follow, we discuss the implementation highlights related 
to the peculiarities of \CEU:
the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

@@

    <...>   // other events
}
\end{lstlisting}
%\rule{14cm}{0.37pt}
\caption{
%{\small %\textmd{
The \emph{TinyOS} binding for \CEU.
This platform-dependent template includes the \C files generated from the 
original application in \CEU (\code{ceu\_app.h} and \code{ceu\_app.c}) for the 
\emph{code generation phase}.
%}%}
\label{lst.impl.tinyos}
}
\end{figure}

@@

A number of virtual machines have been proposed for embedded systems.
%
%The \emph{Sun SPOT} platform with the \emph{Squawk JVM} brings Java for the 
%embedded domain~\cite{wsn.sunspot}, but requires a much powerful hardware.%
%\footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 
%KBytes of SRAM.}
%
\emph{Darjeeling}~\cite{wsn.darjeeling} and \emph{TakaTuka}~\cite{wsn.takatuka} 
are complete \emph{Java VMs} targeting constrained embedded systems with 
support for multithreading and garbage collection.
%
Java has antagonistic design choices in comparison to \CEU:
it does not impose static bounds on memory usage and execution time, and 
provides preemptive multithreading which requires synchronization primitives 
for accessing shared memory.
%
Plummer et al.~\cite{esterel.vm} propose a Esterel-based \VM with similar 
design choices to our work.
To reduce code size, the \VM has a specialized instruction set to deal with 
events and concurrency constructs that are particular to Esterel.
However, the proposed \VM is only a proof of concept, with no support for 
arithmetic operations, external system calls, or remote reprogramming.
